---
title: "Code Library"
author: "Ruhika Chatterjee"
date: "2024-12-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Notes taken from Johns Hopkins University Coursera course series Data Science Specialization.

# R Data Types
- Basic object is vector of same class except list.  
- Atomic classes of objects: character, numeric (real), integer, complex, logical.  
- Attributes can include names, dimnames, dimensions, class, length.

``` {r atomicClass}

# numeric Vector
x <- 5 # numeric vector of 1 element

# integer vector
x <- 5L # integer vector of len 1

x <- Inf # special number infinity, +/-
x <- NaN # special number undefined, usually hijacks operations

# character vector
msg <- "hello" # char vector of len 1

# logical vector
tf <- TRUE # logical vector of value true
# TRUE = 1 = T, FALSE = 0 = F, num > 0 = TRUE

# complex vector
x <- 1+4i # vector of complex num of len 1

```

### complex Data Types
``` {r complexData}
# vector
x <- vector("numeric", length = 10) # create vector of one type, args: class, length
x <- c(1,2,3,4) # creates vector of common denominator class with given values
x <- 1:20 # vector sequence of 20 elements 1-20
x <- pi:10 # will not exceed 10, start from pi, increment by 1
x <- 15:1 # increment -1
x <- seq(1,20) # same as :
x <- seq(0,10,by=0.5) # to change increment
x <- seq(5,10,length=30) # to not set increment but number of numbers
x <- seq_along(x) # vector of same length 1:length(x)
x <- rep(10, times = 4) # repeats 10 4 times in vector
x <- rep(c(0, 1, 2), times = 10) # repeats sequence of vector 10 times. Arg each can be used to repeat first, then second... element in vector.

# vectorized operations
x <- 1:4; y <- 6:9 # different length vectors
x + y # add the elements of the vectors, all operators work
x > 2 # returns logical vector, >= or == or any of the logical expressions work

# lists
# vector capable of carrying different classes
x <- list(1, "a", TRUE, 1+4i) # vector of vectors

# Matrix
# vector of single class with rectangular dimensions (attribute of integer vector len 2)
x <- matrix(nrow=2,ncol=3) # empty matrix of given dimensions
x <- matrix(1:8, nrow = 4, ncol = 2) # creates matrix of given dimensions with values assigned, created by column
y <- matrix(rep(10,4),2,2) # creates matrix of 4 10s

x <- 1:10
dim(x) <- c(2,5) # creates matrix out of vector with dimension 2 rows x 5 columns

cbind(1:3,10:12) # creates matrix out of values in vector args, adding by column (1st arg = 1st col)
rbind(1:3,10:12) # same but using rows

# vectorized operation
x <- matrix(1:4,2,2); y <- matrix(rep(10,4),2,2)
x * y # element wise multiplication, for all operators
x %*% y # matrix multiplication

# factors
# self-describing type of vector representing categorical data, ordered or unordered (labels)
x <- factor(c("male","female","female","female","male")) # character vector with specific linear modeling capabilities, levels also in output
f <- gl(3,10) # factor 3 levels, 10 times each
table(x) # prints counts of each factor

# data frames
# stores tabular/rectangular data, stored as lists of same length where each element is a column, length of element is number of rows. Different classes possible
x <- data.frame(foo=1:4, bar=c(T,T,F,F)) # creates data frame 2 columns foo and bar, 4 rows unnamed. Can also combine with matrices
x <- read.table(file = "hw1_data.csv", header = TRUE, sep = ",") # read in data from file
x <- read.csv("hw1_data.csv") # same
row.names(x) # get and set row names (attributes). Can also use rownames(x)
colnames(x) # get and set row names
nrow(x) # number of rows
ncol(x) # number of columns
data.matrix(x) # converts data frame to matrix, coercion

# names attribute
x <- 1:3
names(x) # is null
names(x) <- c("foo","bar","norf") #now not numbered vector but named, print x and names(x) with names
vect <- c(foo = 11, bar = 2, norf = NA) # adds elements with names to vector directly
# also for lists, names vectors not items
m <- matrix(1:4,nrow = 2, ncol = 2)
dimnames(m) <- list(c("a","b"),c("c","d")) # each dimension has a name for matrices, rows names then columns
```

``` {r datetime}
# useful for time-series data (temporal changes) or other temporal info
# lubridate package by Hadley Wickham

# Dates and Times
birthday <- as.Date("1970-01-01") # dates are date class defined by converting character string, year-month-day, stored as number of days since 1970-01-01 (previous is negative)
today <- Sys.Date()

currentTime <- Sys.time()# time by POSIXct(large integer vector, useful in dataframe) or POSIXlt(list, other info stored like day) class, stored internally as number of seconds since 1970-01-01 or list, respectively
timedefined <- as.POSIXct("2012-10-25 06:00:00") # convert char vector, can define timezone
cTConvert <- as.POSIXlt(currentTime) # reclass, works other way
cTConvert$min # to subset list

datestring <- c("January 10, 2012 10:40", "December 9, 2011 9:10")
x <- strptime(datestring, "%B %d, %Y %H:%M") # Convert character vector to POSIXlt by defining format (format string)
x

weekdays(birthday) # return day of week, date or time classes
months(birthday) # return month on date or time
quarters(birthday) # return quarter of date or time

# Operations
# CANNOT MIX CLASSES - convert
# add and subtract dates, compare dates
currentTime - timedefined # time difference, track of discrepancies (i.e. daylightsavings, timezones, leap time, etc). Can do + and - or <, >, <=, >=, ==, !=.
difftime(currentTime, timedefined, units = "days") # to specify unit

```


# Basic R Functions
``` {r basic}
# managing working directory and work space
x <- getwd() # find working directory
dir.create("testdir") # create a directory, args: dir name, for nested recursive = true
setwd("testdir") # set working dir
file.create("mytest.R") # create file in wd
file.exists("mytest.R") # check if file exists in wd
file.info("mytest.R") # file metadata, use $ operator to grab specific items
file.rename("mytest.R","mytest2.R") # rename
file.copy("mytest2.R","mytest3.R") # copy file
file.remove("mytest2.R") # remove file
file.path("mytest3.R") # relative path
setwd(x)

dir() # output files in directory. Also list.files().
ls() # prints the objects in work space
rm(list=ls()) # clear workspace
rm(list=setdiff(ls(), "x")) # clear workspace except x
version #R info version
sessionInfo() #R info version, packages
source("coded.R") # load code into console
args(ls()) # get arguments for a function
help(ls) # access documentation on ls() function
?ls # same. for operator use ?`:`

# Basic Functions
# add numbers, vectors (element-by-element or recycling)
x <- 5 + 7 # basic arithmetic operations all work +, -, *, /, ^, %% (modulus). NA in expression returns NA.
sqrt(4) # square root
abs(-1:2) # absolute value

# Logical operators
5 >= 2 # returns logical. <, >, <=, >=, ==, !=. NA in expression returns NA. Can also use to compare logical vectors themselves.
TRUE | FALSE # OR A|B union, AND A&B intersection, NOT !A negation. & operates across vector, && evaluates first element of vector. Order of operations: AND before OR
isTRUE(6 > 4) # also evaluates logical expression
xor(5 == 6, !FALSE) # only returns TRUE if one is TRUE, one is FALSE
which(c(1,2,3,4,5,6) < 2) # returns indices of logical vector where element is TRUE
any(c(1,2,3,4,5,6) < 2) # returns TRUE if any of the logical index values are TRUE
all(c(1,2,3,4,5,6) < 2) # returns TRUE only if all the elements of vector are TRUE

# Character functions
paste(c("My","name","is"),collapse = " ") # join elements into one element, can join multiple vectors with multiple arguments, adds element-by-element with coersion and recycling
c (c("My","name","is"), "Bob") # add to the vector

# Input and Evaluation
x <- 1 # assignment operator, evaluates and returns
print(x) # print value as vector
x # auto-prints
# in console, press Tab for auto-completion
LETTERS # predefined character vector of capital letters

# <<- operator can be used to assign a value to an object in an environment that is different from the current environment

# Functions on Objects
class(x) # determine class of object
attributes(x) # function to return or modify attributes of object
identical(x,x) # returns logical for if two objects are identical
length(x) # to specifically get the length of vector
dim(x) # to get dimensions of matrix, data frame (row x column)
object.size(x) # return memory occupied in bytes
c(0.5,0.8,10) # creates a vector of a certain class otherwise coercion
-c(0.5,0.8,10) # distributes the negative to all elements of vector
as.numeric(0:6) # explicit coercion, works on all atomic classes, if not possible converts to NA and warning
mean(c(3,4,5,6,7)) # return mean of numeric vector
sd(c(3,4,5,6,7)) # returns standard deviation of numeric vector
cor(c(3,4,5,6,7), c(61,47,18,18,5)) # correlation of x and y vectors make sure to set arg use for NAs
head(data.frame(foo = 1:20, rar = 301:320)) # prints preview of first 6 lines
tail(data.frame(foo = 1:20, rar = 301:320)) # prints preview of last 6 lines
table(c(1,1,1,2,2,2,2,2,2,2,2,2,3,3,3,3,4,4,5)) # returns table of counts
summary(c(3,4,5,6,7)) # result summaries of the results of various model fitting functions based on class
unique(c(3,4,5,6,7,3,3,5,7,2,8,3,5,6)) # returns only unique elements, duplicates removed
range(c(3,4,5,6,7)) # returns min and max as numeric vector of 2
quantile(c(3,4,5,6,7), probs = 0.25) # returns 25th percentile

# str function - compactly display internal structure of R object (esp large lists). Diagnostic, alternative to summary()
str(unclass(as.POSIXlt(Sys.time()))) # prints list clearly
str(lm) # list of function arguments
str(rnorm(100,2,4)) # type of vector, length, first 5 elements
str(gl(40,10)) # for factors

# Matrix function
x <- matrix(rnorm(200), 20, 10)
rowSums(x) # vector of sum of rows
rowMeans(x) # vector of mean of rows
colSums(x) # vector of sum of cols
colMeans(x) # vector of mean of cols
x <- matrix(rnorm(100), 10, 10)
solve(x) # returns inverse of matrix if invertible

# Factors functions
x <- factor(c("male","female","female","female","male")) # can include levels argument to set order (baseline is first) otherwise alphabetical
x # prints values in vector and levels
table(x) # prints labels and counts present
unclass(x) # strips class to integer with levels of labels

# Missing Values
# represented as NA (missing, with specified class) or NaN (missing or undefined)
# NaN is NA but NA not always NaN
is.na(c (1,2,NA,5,6,NA, NA,3, NaN)) # output logical vector of length of input
is.nan(c (1,2,NaN,5,6,NA, NaN,3)) # output logical vector of length of input

```

``` {r randomnumbers}
# Random number generation
# Probability distribution functions have 4 functions associated: d- density, r- random number generation, p- cumulative distribution, q- quantile function. Normal, gamma, Poisson, binomial, exponential, chi-squared
set.seed(1) # set sequence of random number generation. set.seed(1); rnorm(5) always results in the same sequence (reproducible).
y <- rnorm(1000) # generate vector of 1000 numbers that are standard normal distribution. Agrs: n, mean = 0, sd = 1.
y <- dnorm(c(0.25,0.5,0.75)) # evaluate Normal probability density, (given mean,sd) at point or vector points. Args: x, mean = 0, sd = 1, log = FALSE.
y <- pnorm(0.5) # evaluate cumulative distribution function for normal distribution. Args: q, mean=0, sd=1, lower.tail=TRUE, log.p=FALSE.
y <- qnorm(0.5) # evaluates quantiles for normal distribution. Args: p, mean=0, sd=1, lower.tail=TRUE, log.p=FALSE.
y <- sample(1:6,3) # random selection of 3 elements from array
ints <- sample(10) # random sample all integers from 1 to 10 without replacement. Permutation
nums <- sample(1:10, replace = TRUE) # with replacement
let <- sample(LETTERS) # sample all letters without replacement
flips <- sample(c(0,1), 100, replace = TRUE, prob = c(0.3,0.7)) # unfair coin
coin <- rbinom(1,1,0.5) # simulating coin flip
unfairflip <- rbinom(1, size = 100, prob = 0.7) # sum of flips above
flips2 <- rbinom(100,1,0.7) # flips above
y <- rpois(10, 1) # generate random poisson variates with given rate. Args: n (count), rate (mean)
pois_mat <- replicate(100, rpois(5, 10))

# Simulate Linear Model Ex
# y = B(o) + B(1) * x + e
# e ~ N(0,2^2) assume x ~ N(0,1^2), B(0) = 0.5, B(1) = 2.
set.seed(20)
x <- rnorm(100)
e <- rnorm(100,0,2)
y <- 0.5 + 2 * x + e
# can combine different distributions
# Poisson: Y ~ Poisson(mu)
# log(mu) = B(0) + B(1)x
# B(0) = 0.5 and B(1) = 0.3
set.seed(1)
x <- rnorm(100)
log.mu <- 0.5 + 0.3 * x
y <- rpois(100, exp(log.mu))

```


``` {r controlfunctions}
# control execution of program

x = 2
# if,else loops
y <- if(x > 3){ # testing condition
  10
} else if(x > 0 & x <= 3) { # can not have or multiple
  5
} else{ # can not have, at end
  0
}

if(x-5 == 0){
  y <- 0
} else{
  y <- 2
}

# for loops
for(i in 1:10) {# execute loop fixed number of times. Args iterator variable and vector(inc seq) or list
  print(i)
}

x <- c("a","b","c","d")
for(i in 1:4){
  print(x[i])
}
for(i in seq_along(x)){
  print(x[i])
}
for(letter in x){
  print(letter)
}
for(i in 1:4) print(x[i])

x <- matrix(1:6,2,3)
for(i in seq_len(nrow(x))) { # nested, don't use more than 2-3 for readability
  for(j in seq_len(ncol(x))) {
    print(x[i,j])
  }
}

# while loops
count <- 0
while(count < 10){ # loop while condition is true
  print(count)
  count <- count + 1
} # be wary of infinite loops!! when condition cannot be true

z <- 5
while(z >= 3 & z <= 10){
  print(z)
  coin <- rbinom(1,1,0.5)
  
  if (coin == 1) z <- z+1
  else z <- z-1
}

# Repeat loop
x0 <- 0.01; tol <- 1e-3
repeat { # infinite loop
  x1 <- rnorm(1)
  if(abs(x1 - x0) < tol) {
    break # break execution of any loop
  }
  else x0 <- x1
}

# control a loop
for(i in 1:100) {
  if(i <= 20) next # skip next iteration of loop
  else {
    if (i > 50) break # exit for loop
  }
}

# return to exit a function, will end control structure inside function

```

``` {r loopfunctions}
# Loop functions - useful for looping in the command line
# Hadley Wickham's Journal of Statistical Software paper titled 'The Split-Apply-Combine Strategy for Data Analysis'.

# lapply - loop over a list and evaluate on each element. args: X (list or coercion), FUN (function or function name), ... (other args). returns list with same names if defined. Passes element to function first argument
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean) # returns list of 2 numerics
x <- 1:4
lapply(x, runif, min = 0, max = 10) # passes subsequent args to function
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
lapply(x, function(elt) elt[,1]) # define an anonymous function inside lapply

# sapply - same as lapply but simplify, i.e. will make list of 1 element vectors a vector, multiple elements same length returns a matrix, else a list
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean) # now returns vector length 2
# mean only operates on signle element numeric/logical, so need to use loop

# vapply - pre-specify type of return value, safer and faster. Args: X, FUN, FUN.VALUE (generalized vector as template for return), ..., USE.NAMES = TRUE)
vapply(x, mean, numeric(1)) # same as sapply(x, mean)

# apply - apply function over margins of array (good for summary of matrices or higher level array). Not better than for loop except less code. Args: X as above, MARGIN (integer vector showing margins to retain), FUN as above, ... as above
x <- matrix(rnorm(200), 20, 10)
apply(x, 2, mean) # mean of each column by collapsing 1st dimension, returns num vector length of ncol. MARGIN = 1 gives rows collapse cols.
rowSums(x) # equivalent to apply(x, 1, sum)
rowMeans(x) # equivalent to apply(x, 1, mean)
colSums(x) # apply(x, 2, sum)
colMeans(x) #apply(x, 2, mean)
apply(x, 1, quantile, probs = c(0.25, 0.75)) # runs quantile with 2 agrs for every element in list, returns matrix of 2 rows, nrows rows.
a <- array(rnorm(2 * 2 * 10), c(2, 2, 10)) # array in 3D
apply(a, c(1,2), mean) # collapses only 3rd dimension, returns 2x2 matrix. Equivalent rowMeans(a, dims = 2).

# tapply - apply function over subset of a vector. args: X is vector, INDEX is factor/list factors vector same length as X, FUN, ..., simplify = TRUE
x <- c(rnorm(10), runif(10), rnorm(10,1))
f <- gl(3,10) # factor 3 levels, 10 times each
tapply(x,f,mean)

# mapply - multivariate version of lapply. args: FUN as above, ... (arguments to apply over), MoreArgs (other arguments for FUN), SIMPLIFY (logical for if to simplify). Apply over multiple lists. Like nested loop
list(rep(1,4), rep(2,3), rep(3,2), rep(4,1))
mapply(rep, 1:4, 4:1) # equivalent

noise <- function(n,mean,sd){rnorm(n,mean,sd)}
noise(1:5,1:5,2) # gives vector of 5, same as single num args
mapply(noise,1:5,1:5,2) # applies function for each pair, list of 5 of length i

# split - in conjunction with lapply to split objects into subpieces. Args: x (any object), f (factor), drop (logical if empty factors to be dropped) = FALSE, ...
x <- c(rnorm(10), runif(10), rnorm(10,1))
f <- gl(3,10) # factor 3 levels, 10 times each
split(x,f) # tapply without function, sorts into list based on levels, can then use lapply or sapply.
lapply(split(x,f), mean) # in this case can use tapply
# can do data frames
data <- read.csv("hw1_data.csv")
s <- split(data, data$Month)
sapply(s, function(x) colMeans(x[,c("Ozone","Solar.R","Wind")], na.rm = TRUE)) # data$Month coerced into factor
# Multi-level split
x <- rnorm(10)
f1 <- gl(2,5); f2 <- gl(5,2) # ex. race and gender 2 factors
interaction(f1,f1) # combine each pair, 10 factors
split(x, list(f1,f2)) # interaction called, list returned for combination sort, drop = TRUE to remove unused

```

``` {r defFunctions}
# stored in txt or R script, functions are R objects. Can pass functions as arguments for other functions, functions can be nested.

myfunction <- function(){ #create a function
  x <- rnorm(100)
  mean(x)
}
myfunction() #call created function
myfunction # prints source code for function
args(myfunction) # returns arguments for passed function

myaddedfunction <- function(x,y){ #create a function with formal arguments x and y
  x + y + rnorm(100) # implicit return last expression
}
myaddedfunction(5,3)
myaddedfunction(4:10,2)

# function with default argument if left unspecified, for common cases
above <- function(x, n = 10){
  use <- x > n
  x[use]
}
above(1:20) # n is default set to 10
above(1:20, 12) # n set at 12

columnmean <- function(y, removeNA = TRUE) {
  nc <- ncol(y)
  means <- numeric(nc)
  for(i in 1:nc) means[i] <- mean(y[,i], na.rm = removeNA)
  invisible(means) # auto-return blocks auto-print
}

# Lazy Evaluation: R evaluated statements and arguments as they come
f <- function (a,b,c){
  print(a)
  #print(b) # error
}
f(3) # prints a, error for b, no rxn to not having c

# ways to call functions
# positional matching and naming can be mixed. Partial matching also allowed, if not found uses positional match.
# named helps for long arg list where most defaults are maintained or if order is hard to remember.
mydata <- rnorm(100)
sd(mydata) # default to first argument
sd(x = mydata)
sd(x = mydata, na.rm = FALSE)
sd(na.rm = FALSE, x = mydata)
sd(na.rm = FALSE, mydata) # remove argument from list, default works on first unspecified arg

# Variable Arguments
# to extend another function without copying arg list of OG function
simon_says <- function(...){
  paste("Simon says:", ...)
}
# or for generic functions passed to methods
# unpacking an ellipses
mad_libs <- function(...){
  args <- list(...)
  place <- args$place
  adjective <- args$adjective
  noun <- args$noun
  paste("News from", place, "today where", adjective, "students took to the streets in protest of the new", noun, "being installed on campus.")
}
# or when number of args unknown in advance (if at beginning, no positional or partial matching)
args(paste) # operates on unknown sets of character vectors

# function as an argument
some_function <- function(func){
  func(2, 4) # returns result of function with 2,4 arguments
}
some_function(mean) # returns mean of 2,4

# Anonymous function (chaos)
evaluate  <- function(func, dat){
  func(dat)
}
evaluate(function(x){x+1}, 6) # creates a function when calling evaluate to add 1

# create a binary operation
"%mult_add_one%" <- function(left, right){
  left * right + 1
}
4 %mult_add_one% 5

```

``` {r lexicalscoping}
make.power <- function(n) {
  pow <- function(x) {
    x^n
  }
  pow
}

cube <- make.power(3)
square <- make.power(2)
cube(3)
square(3)

# Scoping - environments
search()# provides list of environments
ls(environment(cube)) # object names in function environment, same for square
get("n",environment(cube)) # values in function environment, changes for square

```


# R Packages
- Repositories: CRAN, BioConductor (bioinformatics), GitHub  
- Search: https://www.rdocumentation.org/  

- Base packages: utils, stats, datasets, graphics, grDevices, grid, methods, tools, parallel, compiler, splines, tcltk, stats4.  
- Recommended packages: boot, class, cluster, codetools, foreign, KernSmooth, lattice, mgcv, nime, rpart, survival, MASS, spatial, nnet, Matrix.  
``` {r packages}
# Install from CRAN:
#   install.packages("ggplot2", repos = "http://cran.us.r-project.org") #install
#   install.packages(c("labeling","tibble"), repos = "http://cran.us.r-project.org") #multiple

# Install from Bioconductor
#   install.packages("BiocManager", repos = "https://bioconductor.org/biocLite.R")
#   BiocManager::install(c("GenomicFeatures", "AnnotationDbi")) #install package

# Install from GitHub (need package, author name)
#   install.packages("devtools", repos = "http://cran.us.r-project.org") #only once
#   library(devtools)
#   install_github("author/package") #installs package

# library(ggplot2)# Load package, careful of dependencies
# installed.packages() #check installed packages
# library() #alternate
# old.packages(repos = "http://cran.us.r-project.org") #check packages to update
# update.packages(repos = "http://cran.us.r-project.org") #update all packages
# install.packages("ggplot2") #to update single package
# detach("package:ggplot2", unload=TRUE) #unload function
# remove.packages("ggtree") #remove package
# help(package = "ggplot2") #package info
# browseVignettes("ggplot2") #extended help files

```


# Data in R

``` {r workingData}

# Pull all file names from a directory into a character vector
files_full <- list.files("specdata", full.names=TRUE)

# Read data into R
x <- read.table("hw1_data.csv", header = TRUE, sep = ",") #reading tabular data from text files, return data frame. Args: file (name, connection), header (logical if header line present), sep (string indicating column separator), colClasses (character vector of class of each column), nrows (number of rows), comment.char (character string indicating comment character), skip (number of lines to skip from beginning), stringsAsFactors (logical if character variables coded as factors, default true). Efficient. Default separator " " and header = FALSE. Check help page to optimize for large datasets, set comment.char = "" to optimize if no comments present, use colClasses, nrow (memory).
x <- read.csv("hw1_data.csv") # Same but default separator is ", "  and header = TRUE
write.table(x)

# help read.table with colClasses with smaller sample
initial <- read.table("hw1_data.csv", header = TRUE, sep = ",", nrows = 100)
classes <- sapply(initial, class)
tabAll <- read.table("hw1_data.csv", header = TRUE, sep = ",", colClasses = classes)

lines <- readLines("coded.R") # reading lines of text file, return character vector
writeLines("coded.R")

# editable textual format retains metadata, helpful for version control, corruption fixable, memory cost
dget("coded.R") # reading R objects deparsed into text files
dput("coded.R") # takes R object, create R code to reconstruct object saving attributes, names
source("coded.R") # reading in R code files
#dump() # multiple R objects

# load() # read in saved workspace read binary objects into R
# save()
# unserialize() # read single R objects in binary form
# serialize()

# Interface to outside world
file(description = "hw1_data.csv") # open connection to standard, uncompressed file. Helps for partial file reading. description = file name, open = "r"(read only), "w" (writing and initializing new file), "a" (appending), "rb"/"wb"/"ab" in binary (Windows)
# gzfile() # connection to file w compression gzip
# bzfile() # connection to file w compression bzip2
jh <- url("http://www.jhsph.edu", "r") # connection to webpage
close(jh) # to end connection

# Subsetting R Objects
x <- c("a","b","c","c","d","a")
x[1] # more than one element extracted, returns same class as the original, numeric/logical index
x[1:4] # sequence of num index
x[x>"a"] # logical indexing, returns vector where logical is true
u <- x > "a" # create logical vector
x[u] # same as x[x>"a"]
x[!is.na(x) & x > 0] # returns only positive, non NA values
x[c(-2, -10)] # returns vector with 2nd and 10th elements removed
x <- data.frame(foo = 1:6, bar = c("g","h","i","j","k","l"))
x[[which(x$bar == "h"), "foo"]] # get or set foo in the same row as bar of "h"

x <- list(foo = 1:4, bar = 0.6, baz = "hello")
x[1] # list containing first element
x[[1]] # extract from list/data frame, single element, class can change. Ex, numerical vector returned
x$bar # like [[]] but by name. Ex, return num vector 0.6. Equivalent to x[["bar"]]. Expression x["bar"] returns list of bar. Can use partial matching (x$f = x$foo = x[["f", exact = FALSE]]).
x[c(1,3)] # multiple object extraction from list, returns list
name = "foo"
x[[name]] # must be used if using computed index
x[1][3] # return element in element in object
x[[c(1,3)]]

# Subsetting Matrix
x <- matrix(1:6, 2, 3)
x [1,2] # returns vector len 1, different that x[2,1]. Get matrix using arg drop = FALSE.
x[1,] # get num vector of first row, can also get col x[,2]. drop = FALSE also works

# Removing NA values
x <- c(1,2,NA,4,NA,5)
bad <- is.na(x) # logical vector indicating presence of NA
x[!bad] # removes NA values

x[!is.na(x)] # simplified returns vector removing NA values

x <- c(1,2,NA,4,NA,5) # for two vectors
y <- c("a","b",NA,"d",NA,"f")
good <- complete.cases(x,y) # logical vectors where there is no NA in either list
x[good]
y[good]

# Sum of NA values
my_na <- is.na(x)
sum(my_na)

x <- read.csv("hw1_data.csv") # for data frames
goodVals <- complete.cases(x) # complete rows in the data frame
x[goodVals,]

```

# Graphics
<http://www.ling.upenn.edu/~joseff/rstudy/week4.html>
``` {r basegraphs}
# Start by getting sense of the data: dim(), names(), head(), tail() and summary().

# Scatterplot
data(cars)
plot(cars) # generates scatterplot with two columns against each other
plot(x = cars$speed, y = cars$dist) # explicit, can also pass plot(dist ~ speed, cars). Args: x, y = NULL, type = "p",  xlim = NULL, ylim = NULL, log = "", main = NULL, sub = NULL, xlab = NULL, ylab = NULL, ann = par("ann"), axes = TRUE, frame.plot = axes, panel.first = NULL, panel.last = NULL, asp = NA, xgap.axis = NA, ygap.axis = NA, ... (more at ?par)

# Boxplot
data(mtcars)
boxplot(mpg ~ cyl, data = mtcars) # generate relationship between mpg(x) and cyl(y) from mtcars

# Histogram
hist(mtcars$mpg) # generate a histogram of vector

```

# R Profiler and Optimization
- Systematic way to examine time spent in various part of the program. Useful to optimize the code.  
- DON'T PREMATURELY OPTIMIZE  
- Measure, not guess, data on what needs to be optimized.  
- User time: computer experienced, may be greater if multiple cores/processors (accessible in multi-threaded BLAS libraries). Elapsed time: wall-clock time, may be greater if other computing tasks.  
``` {r profiler}
system.time(read.csv("hw1_data.csv")) # returns seconds to execute, if error then seconds to error. Wrap longer expression in arguement in curly braces. Assumption htat you know where the problem is.

data(mtcars)
Rprof() # track function call stack at intervals (def = 0.02 sec), time spent in functions.

fit <- lm(mtcars$mpg ~ mtcars$cyl)

Rprof(NULL)

summaryRprof() # makes Rprof readable, tabluates, time in each function
# $by.total - divides time spent per function by total run time
# $by.self - same as by.total but first subtracts time spent in function above in call stack. Helps target.

```
